Retrofitting Effect Handlers onto OCaml

### 2020-11-12 retro-concurrency.tex
add note about GOMAXPROCS
include go version number
Updates

\vspace{-5mm}
\subsection{Dynamic semantics}

@@

\vspace{-2mm}
\subsection{Heap-allocated fibers}

@@

\vspace{-3mm}
\subsection{External calls and callbacks}

@@

\vspace{-2mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

\vspace{-2mm}
\subsection{Stack unwinding}

@@

\vspace{-2mm}
\subsection{Garbage collection}

@@

\vspace{-3mm}
\section{Evaluation}
\label{sec:eval}

@@

\textbf{\textit{Web server.}} As one of our primary goals of adding effect handlers to
OCaml is to support direct-style concurrency, we implement a full-fledged
HTTP/1.1 web server using effect handlers (|MC|). We use effect handlers to
implement a user-level threading library, which spawns a thread per request. We
use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, Lwt version is written in monadic
style, and does not have a notion of thread per request. For comparison, we
also include a Go version (|go|) that uses \texttt{net/http}~\cite{nethttp}
package. As both OCaml versions are single threaded, the Go benchmark is run with
\texttt{GOMAXPROCS=1}.

@@

In addition, multi-shot continuations currently are not safe under compiler
optimisations. For example consider the following program:
\begin{lstlisting}
effect E : unit

let foo () = perform E

let bar f =
  let r = ref 0 in
	f ();
  r := !r + 1;
  !r

let _ = match bar foo with
| v -> v
| effect E k ->
    let k' = Obj.clone_continuation k in
    let r1 = continue k () in
    let r2 = continue k' () in
    Printf.printf "%d %d\n" r1 r2;
		0
\end{lstlisting}
This program is expected to print |1 2|. However, OCaml optimizes |bar| by
converting the heap allocation of |r| to a stack allocation. As a result, each
continuation ends up with a copy of |r|, and the program prints |1 1|. In the
presence of multi-shot continuations, compiler optimisations need to be made
aware of the effects of a function. We leave proper support for multi-shot
continuations to future work.

>>>>

\subsection{Dynamic semantics}

@@

\subsection{Heap-allocated fibers}

@@

\subsection{External calls and callbacks}

@@

\subsection{Effect handlers}
\label{sec:effimpl}

@@

\vspace{-2.5mm}
\subsection{Stack unwinding}

@@

\subsection{Garbage collection}

@@

\section{Evaluation}
\label{sec:eval}

@@

\textbf{\textit{Web server.}} As one of our primary goals of adding effect
handlers to OCaml is to support direct-style concurrency, we implement a
full-fledged HTTP/1.1 web server using effect handlers (|MC|). We use effect
handlers to implement a user-level threading library, which spawns a thread per
request. We use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, Lwt version is written in monadic
style, and does not have a notion of thread per request. For comparison, we
also include a Go 1.13 version (|go|) that uses
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.

@@

In addition, multi-shot continuations currently are not safe under compiler
optimisations. For example consider the following program:

\begin{minipage}{\linewidth}
\begin{lstlisting}
effect E : unit
let foo () = perform E
let bar f =
  let r = ref 0 in
	f ();
  r := !r + 1;
  !r
;;
match bar foo with
| v -> v
| effect E k ->
    let k' = Obj.clone_continuation k in
    let r1 = continue k () in
    let r2 = continue k' () in
    Printf.printf "%d %d\n" r1 r2;
		0
\end{lstlisting}
\end{minipage}

This program is expected to print |1 2|. However, OCaml optimizes |bar| by
converting the heap allocation of |r| to a stack allocation. As a result, each
continuation ends up with a copy of |r|, and the program prints |1 1|. Thus, in
the presence of multi-shot continuations, compiler optimisations need to be
made aware of the effects of a function. We leave proper support for multi-shot
continuations to future work.
